Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-26 submission/manuscript.Rmd
shigeki results 1 edits

Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017]. Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, it is currently unclear if this erosion of brain tissue identity is a major consequence of _Kdm5c_ loss and if it is unique to testis-enriched genes. Therefore, we first globally assessed the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the adult amygdala and hippocampus in mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].

@@

We found a large proportion of genes that are significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched in non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). The majority of tissue-enriched differentially expressed genes (tissue-enriched DEGs) were testis genes (Figure 1A-C). Even though the testis has the largest total number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). 
One example of a testis-enriched DEG is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

In addition to the high enrichment of testis genes, we also identified aberrant expression of other tissue-enriched genes within the _Kdm5c_-KO brain. Although the _Kdm5c_-KO mice were male, we observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1D). Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which was recently shown to sequester mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D). Although not consistent across brain regions, we also found significant enrichment of DEGS biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test). An example of a liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, which is involved in lipoprotein metabolism (Figure 1E). Testis, ovary, and liver-enriched DEGs showed little to no expression in the developing and adult wild-type brain, yet our mRNA-seq data indicate they are polyadenylated and spliced into mature transcripts (Figure 1C-E). Of note, we did not observe enrichment of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes). Together, these results suggest misexpression of tissue-enriched genes within the brain is a major effect of KDM5C loss.

>>>>

Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017]. Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, if the testis is the only tissue type misexpressed and if loss of tissue identity a major consequence of KDM5C loss. Therefore, we globally assessed the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the adult amygdala and hippocampus in mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].

@@

We found a large proportion of genes that are significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are enriched in non-brain tissues in wild-type mice (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). The majority of tissue-enriched differentially expressed genes (tissue-enriched DEGs) were testis genes (Figure 1A-C). Even though the testis has the largest total number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). One example of a testis-enriched DEG is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

In addition to the high enrichment of testis genes, we also identified aberrant expression of other tissue-enriched genes within the _Kdm5c_-KO brain. Although the _Kdm5c_-KO mice we examined were male, we observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1D). Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, sequesters mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D). Although not consistent across brain regions, we also found significant enrichment of DEGS biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test). An example of a liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] (Figure 1E). Testis, ovary, and liver-enriched DEGs showed little to no expression in the developing and adult wild-type brain, yet our mRNA-seq data indicate they are polyadenylated and spliced into mature transcripts (Figure 1C-E). Of note, we did not observe enrichment of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes). Together, these results suggest misexpression of tissue-enriched genes within the brain is a major effect of KDM5C loss.
